A primer to numerical simulations: The perihelion motion of Mercury

### 2018-04-17 ped-response/response.tex

%Bira 06/04/11
%Rob 06/04/11
\documentclass[prc,preprint,showpacs,showkeys,nofootinbib,tightenlines]{revtex4}
\newcommand{\slashT}{\slash\hspace{-0.4em}T}

\usepackage{amsmath}

\begin{document}
\pagenumbering{gobble}

\section*{Letter to the Editor}

\noindent
Dear Editor, \\

We thank the referee for the positive and constructive report.
As detailed below we followed the suggestions made by the referee and hope that the improved version is now regarded appropriate for publication in Physics Education.

\begin{enumerate}
\item \textbf{Referee:} 
\textit{
Section 2 is largely unnecessary as it covers the basics of Newton's Laws and the Euler step method, which will be familiar to everyone.
This section could be shortened to the necessary equations and an introduction to the notation, together with a comment on time steps as this is important later.
}
\\
\\
\textbf{Response:}
We shortened the section somewhat, however, kept most of the equations. In particular, although straightforward, we regard it as useful for the logic of the presentation (and probably also for teaching the course) to introduce the derivatives via Eqs (1) and (2) and not directly via Eqs (3) and (4).

\item \textbf{Referee:} 
\textit{
figure 2 shows that the trajectory varying even when alpha and beta are zero.
It is worth explaining at this point (rather than later) that this is due to the finite time step.
Also it is worth noting if the trajectory change from one orbit to the next leads to a perihelion shift (also picked up later in more detail)
}
\\
\\
\textbf{Response:}
We moved half of Fig. 2 forward into Sec. 2. We thank the referee for this suggestion for it allows us to illustrate the statements immediately and quantitatively, which improves the presentation significantly.

\item \textbf{Referee:} 
\textit{
Finally, the only reason this paper is suitable for Phys Ed rather than, say a journal aimed at undergraduates, is that the authors claim it has been used successfully on students of high school age.
However, there is no supporting evidence of this, no indication of success criteria and no explanation of what was done.
In order to fit better with the idea of an education research journal at this level, perhaps the authors could include some feedback or anecdotal evidence even if they have no formal study of the efficacy of the trial?
}
\\
\\
\textbf{Response:}
We added the new Sec. 8 (Own experiences with the implementation of the course), where we now not only explain in which context we used (and will be using) the course, but also provide some information about the participants.

\end{enumerate}

\noindent
Sincerely,\\

	C.~K\"orber,
	I.~Hammer,
	J.-L.~Wynen,
	J.~Heuer,
	C.~M\"uller and
	C.~Hanhart
\newpage

\end{document}

### 2018-04-17 tex/MercuryNumerical.tex
Created latex version of response and adjusted main tex file
Reincluded figure of alpha=0 in sec 4. And made modifications to sec 8.

Numerical simulations play a key role in modern physics because they allow one to tackle some theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep understanding of Differential Geometry. For the course presented here, however, very little math is necessary, such that high school students vaguely familiar to vector calculus and derivatives will benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017. As an example on how the course can be implemented in practice we describe our own
experiences with it in the second to last section.

@@

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in Figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also Figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing either $\alpha\neq0$ or $\beta\neq0$
[Figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ or $\beta$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ or $\beta$ larger than $10^5$ are necessary to get a visible effect.
For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [Figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(b)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}

@@

The idea that a numerical study of the perihelion motion of Mercury is an interesting subject for high
school students was taken from a project work of one of the authors (JH) presented at the 
M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014. Based on this study
the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} to be used in 
the Sch\"ulerakademie Teilchenphysik" in 2015, which aims at  high school students from 10$^{\rm th}$ to 
13$^{\rm th}$ grade. This academy is financed by the German Research Society (DFG) as the outreach
branch
of a large research grant focussing on basic research in particle and nuclear physics.
One of the authors (CH) is acting as a Principal Investigator for the outreach
activities within this scheme. It runs biannually
at the Science Center Overbach in J\"ulich
for four days and has about 25 participants coming from various parts of Germany. 
The academy comprises lectures, a tour to the particle accelerator
COSY as well as the high performance computing facility at the Research Center J\"ulich as well
as one full day of practical work. The course presented here
 is offered as one option out of three on this day and was prepared
by a lecture to the whole group explaining the basics of numerical simulations. 
Given the setting the participating students
are highly motivated and eager to understand. Typically about one half of the participants chose to
work on the numerical project --- out of those about half had programmed a computer before while
the other half did not yet have any programming experience. Most of the participants had already
learned about derivatives in school. Given the very positive to enthusiastic 
feed backs we received from the participants we decided to make the course in the form presented
to a permanent part of the academy. It was used in the 2017 with the same success.

The students worked in pairs. Each group had an own laptop provided by us, where the necessary
software was preinstalled. Four of the authors (IH, CK, CM, JLW) acted as supervisors for
the students. 
The material covered in this one day is in most cases what is presented in Sec.~\ref{sec:Numerical Implementation}.
Only very few participants managed to start thinking about how to actually extract quantitatively the
perihelion motion --- not to speak of uncertainty estimates.
Once the basic problem was tackled some students deviated from the suggested path and for
instance studied the problem with one additional planet which was fine for us as well. This is
a very interesting problem by itself for there are parameter ranges where the classical three-body
system develops chaotic features.
 The reason why we included the
additional material in the paper is that we wanted to provide teachers with sufficient material to
probably even fill a working group outside the general curriculum running for a longer period.

>>>>

Numerical simulations play a key role in modern physics because they allow one to tackle some theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep understanding of Differential Geometry. For the course presented here, however, very little math is necessary, such that high school students vaguely familiar to vector calculus and derivatives will benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017. As an example on how the course can be implemented in practice, we describe our own
experiences with it in the second to last section.

@@

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in Figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also Figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing either $\alpha\neq0$ or $\beta\neq0$
[Figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ or $\beta$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ or $\beta$ larger than $10^5$ are necessary to get a visible effect.
For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [Figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=0 $ and $\Delta t = \Delta t_0 / 20$;
		\item[(b)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}

@@

The idea that the numerical study of the perihelion motion of Mercury is an interesting subject for high school students was taken from a project work of one of the authors (JH).
The project was eventually presented at the M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014.
Based on this study, the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} and used in the Sch\"ulerakademie Teilchenphysik in 2015, which aims at high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
This academy is financed by the German Research Society (DFG) as the outreach branch of a research grant focussing on basic research in particle and nuclear physics.
One of the authors (CH) is acting as a Principal Investigator for the outreach activities within this scheme.
The Sch\"ulerakademie Teilchenphysik runs biannually at the Science Center Overbach in J\"ulich for four days and has about 25 participants coming from various parts of Germany.
The academy comprises lectures, a tour to the particle accelerator COSY as well as the high performance computing facility at the Research Center J\"ulich as well as one full day of practical work.
Given the setting the participating students are highly motivated and eager to understand.

The course presented in this paper is offered as one out of three hands-on project options and was prepared by a lecture to the whole group explaining the basics of numerical simulations.
Typically about one half of the participants chose to work on the numerical project --- out of those about one half had previous programming experience.
Most of the participants had already learned about derivatives and vector calculus in school.
Given the positive to enthusiastic feedback we have received from the participants on the numerical course, we decided to make the course a permanent part of the academy.
It was used in the 2017 with the same success.

The numerical course was allocated for one work day (seven hours) and the students worked in pairs.
Each group worked on a laptop provided by us, where the necessary software was preinstalled.
Four of the authors (IH, CK, CM, JLW) acted as supervisors for the students. 
The course was initiated with a brief summary of previously presented methods, motivations and objectives of the simulation.
In the first two hours the students were introduced to \python{} and \vpython{}.
In the following part the students worked with an initial code template and were motivated to come up with own solutions to fulfill the objectives.
A large part of the material covered during this day is presented in Sec.~\ref{sec:Numerical Implementation}.
While a majority of the participants where able to visually simulate the motion of mercury including forces from GR, only a few participants started to implement the quantitative extraction of the perihelion motion --- not to speak of uncertainty estimates.
Once the basic problem was tackled some students deviated from the suggested path and for instance studied the problem with one additional planet or different starting parameters.
This is a very interesting problem by itself for there are parameter ranges where the classical three-body system develops chaotic features.
We encouraged such explorations as well.
The reason why we included the additional material in the paper is that we wanted to provide teachers with sufficient material to probably even fill a working group outside the general curriculum running for a longer period.
